---
title: "Predicting Loan Status"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
###Load Data, Preprocessing
```{r}
setwd("~/Documents/DS 501/Data")
loan <- read.csv('loan.csv', stringsAsFactors = FALSE)
loanT <- loan
num.NA <- sort(sapply(loan, function(x){sum(is.na(x)|x=='' )}), decreasing = TRUE)
remain.col <- names(num.NA) [which(num.NA <= 0.8*dim(loan) [1])]
loan <- loan[, remain.col]
```

### Processing dependent variable
```{r}
sort(table(loan$loan_status))
loan$loan_status <- gsub('Does not meet the credit policy. Status:',
                         '', loan$loan_status)
sort(table(loan$loan_status))
loan <- subset(loan, !loan_status %in% c('Current', 'Issued'))
loan$loan_status_binary <- with(loan, ifelse(loan_status == 'Fully Paid', 1, 0))
```
### Pre-processing user(general) feature
```{r}
# Remove irrelevant features: addr_state, emp_title, member_id, zipcode
# Remaining features: emp_length, home_ownership
loan$home_ownership <- ifelse(loan$home_ownership %in% c('ANY', 'NONE', 'OTHER'), 'OTHER',
                              loan$home_ownership)
int_state <- by(loan, loan$addr_state, function(x) {
  return(mean(x$int_rate))
})
loan$state_mean_int <-
  ifelse(loan$addr_state %in% names(int_state)[which(int_state <=
                                                       quantile(int_state, 0.25))], 'low',
         ifelse(loan$addr_state %in% names(int_state)[which(int_state <=
                                                              quantile(int_state, 0.5))],'lowmedium',
                ifelse(loan$addr_state %in% names(int_state)[which(int_state <= quantile(int_state, 0.75))], 
                       'mediumhigh', 'high')))
select.features_1 <- c("home_ownership", "state_mean_int")
```
### Pre-processing user financial feature
```{r}
loan$annual_inc[which(is.na(loan$annual_inc))] <- median(loan$annual_inc, na.rm = T)
loan$log_annual_inc <- log(loan$annual_inc + 1)
# Verification status may not be available, so not included in the model
select.features_2 <- c("dti", "annual_inc")
```
###Pre-processing credit lines feature
```{r}
# many of the features were removed due to large number of missing values. Missing values of the following reserved features are dealt:
loan$tot_cur_bal[which(is.na(loan$tot_cur_bal))] <- median(loan$tot_cur_bal, na.rm = T)
loan$total_acc[which(is.na(loan$total_acc))] <- median(loan$total_acc, na.rm = T)
loan$open_acc[which(is.na(loan$open_acc))] <- median(loan$open_acc, na.rm = T)
loan$acc_now_delinq[which(is.na(loan$acc_now_delinq))] <- median(loan$acc_now_delinq, na.rm = T)
loan$delinq_2yrs[which(is.na(loan$delinq_2yrs))] <- median(loan$delinq_2yrs, na.rm = T)
loan$mths_since_last_delinq[which(is.na(loan$mths_since_last_delinq))] <- median(loan$mths_since_last_delinq, na.rm = T)
loan$collections_12_mths_ex_med[which(is.na(loan$collections_12_mths_ex_med))] <- median(loan$collections_12_mths_ex_med, na.rm = T)
loan$tot_coll_amt[which(is.na(loan$tot_coll_amt))] <- median(loan$tot_coll_amt, na.rm = T)
loan$pub_rec[which(is.na(loan$pub_rec))] <- median(loan$pub_rec, na.rm = T)
loan$mths_since_last_major_derog[which(is.na(loan$mths_since_last_major_derog))] <- median(loan$mths_since_last_major_derog, na.rm = T)
loan$revol_util[which(is.na(loan$revol_util))] <- median(loan$revol_util, na.rm = T)
loan$total_rev_hi_lim[which(is.na(loan$total_rev_hi_lim))] <- median(loan$total_rev_hi_lim, na.rm = T)
select.features_3 <- c("tot_cur_bal","total_acc","open_acc","acc_now_delinq","delinq_2yrs","mths_since_last_delinq","collections_12_mths_ex_med","tot_coll_amt","pub_rec","mths_since_last_major_derog","revol_util","total_rev_hi_lim")
```
###Pre-processing loan feature & loan payment feature
```{r}
# Irrelevant and unavailable features were removed. Following features were reserved
select.features_4 <- c("loan_amnt","term","installment","funded_amnt","pymnt_plan","recoveries","collection_recovery_fee","out_prncp","out_prncp_inv")
```
### Outcome variables (e.g., grade, subgrade, interest rate) are unavailable, thus not included in the model
```{r}
select.features <- c(select.features_1,select.features_2,select.features_3,select.features_4,"loan_status_binary")
loan <- loan[select.features]
```

### Split train and test dataset
```{r}
train.ind <- sample(1:dim(loan)[1], 0.7 * dim(loan)[1])
train.sub <- loan[train.ind, select.features]
train.sub$state_mean_int <- relevel(as.factor(train.sub$state_mean_int), ref = 'low')
test <- loan[-train.ind, select.features]
```
### Model Fit
```{r}
logis.mod <- glm(loan_status_binary ~ ., train.sub, family = 'binomial')
summary(logis.mod)
```
### Model Evaluation
```{r}
library(pROC)
pred <- predict(logis.mod, test)
plot.roc(test$loan_status_binary, pred)
```
### Let's add regularization
```{r}
library(glmnet)
test.matrix <- model.matrix(~., test[, 1:25])
ind <- train.sub[, 1:25]
ind <- model.matrix( ~., ind)
dep <- train.sub[, "loan_status_binary"]
logis.cvfit <- cv.glmnet(ind, dep, family = 'binomial')
plot(logis.cvfit)
```

```{r}
cv.pred <- predict(logis.cvfit, s = logis.cvfit$lambda.1se, newx = test.matrix)
plot.roc(test$loan_status_binary, cv.pred)
```
### Coefficients of this model
```{r}
print(coef(logis.cvfit, s= "lambda.1se"))
```
```{r}
summary(logis.cvfit)
```








