---
title: "Qian_Zhou_Lending_Club"
output: pdf_document
---

### Step 1: Research Question: What model/which features best predict interest rate?

## 1.1 Understanding Data & Pre-processing
```{r}
loan <- read.csv('loan.csv', stringsAsFactors = FALSE)
loanT <- loan
head(loan)
```
```{r}
#check dimension
print(dim(loan))

#check data format
str(loan)
```
```{r}
# removing columns/features with over 80% NA values
num.NA <- sort(sapply(loan, function(x){sum(is.na(x)|x=='' )}), decreasing = TRUE)
remain.col <- names(num.NA) [which(num.NA <= 0.8*dim(loan) [1])]
loan <- loan[, remain.col]
remain.col

# check remaining feature types
loan.col.type <- sapply(loan, function(x) {typeof(x)})
table(loan.col.type)
```

### Step 2: EDA: Which features influence interest rate? 
```{r}
#Understanding dependent variable
loan$annual_inc[which(is.na(loan$annual_inc))] <- median(loan$annual_inc, na.rm = T)
mean(loan$int_rate)
sd(loan$int_rate)
median(loan$int_rate)
quantile(loan$int_rate, c(0.1, 0.25, 0.5, 0.75, 0.9))
plot(density(loan$int_rate))
```
## 2.1 Exploring Numerical Features
```{r}
# Correlation between int_rate and other numerical features
library(corrplot)
num.feature <- names(loan[, sapply(loan, is.numeric)])
correlation <- cor(loan$int_rate, loan[, c(num.feature)], use = "pairwise.complete.obs")
correlation <- correlation[, order(correlation[1,], decreasing = TRUE)] 
correlation <- correlation[2:33]
#Correlation Plot
corrplot(correlations, method = "square", t1.cex = 1, type = "lower")
```

```{r}
#Correlation between int_rate & top five numerical features
{barplot(correlation, main = 'Correlation between int_rate with other numerical features', ylab = 'Correlation', las=2,
        cex.names=0.6, col=ifelse(correlation>0.15,"red","blue"))
legend("topright", leg = c('top five correlated feature', 'other numerical feature'), fill = c("red", "blue"))}
```
## Based on correlation calculation, the top five predictive numerical features for int_rate are "total_rec_int", "revol_util", "inq_last_6mths", "total_pymnt_inv" and "total_pymnt". 

## 2.2 Exploring Categorical Features
```{r}
#Initial Exploration: Boxplot
library(lattice)
bwplot(int_rate ~ loan_status, data = loan, xlab='loan_status', ylab='int_rat', main='loan_status vs int_rate')
bwplot(int_rate ~ grade, data = loan, xlab='grade', ylab='int_rat', main='grade vs int_rate')
bwplot(int_rate ~ sub_grade, data = loan, xlab='sub_grade', ylab='int_rat', main='sub_grade vs int_rate')
bwplot(int_rate ~ home_ownership, data = loan, xlab='home_ownership', ylab='int_rat', main='home_ownership vs int_rate')
bwplot(int_rate ~ term, data = loan, xlab='term', ylab='int_rat', main='term vs int_rate')
bwplot(int_rate ~ pymnt_plan, data = loan, xlab='pymnt_plan', ylab='int_rat', main='pymnt_plan vs int_rate')
bwplot(int_rate ~ application_type, data = loan, xlab='application_type', ylab='int_rat', main='application_type vs int_rate')
bwplot(int_rate ~ emp_length, data = loan, xlab='emp_length', ylab='int_rat', main='emp_length vs int_rate')
bwplot(int_rate ~ verification_status, data = loan, xlab='verification_status', ylab='int_rat', main='verification_status vs int_rate')
```
## Based on the above ouput, six categorical features seem to be influential: "grade", "sub-grade", "term", "verification_status", "pymnt_plan", "application_type". However, "grade" and "sub-grade" are assigned by LC, thus should not be available at the beginning. Let's test the remaining four variables.
```{r}
# Testing categorical features
#verification_status
t.test(subset(loan, verification_status == 'Verified')$int_rate, subset(loan, verification_status == 'Source Verified')$int_rate, conf.level = 0.95, mu = 0, alternative = "two.sided", paired = FALSE, var.equal = FALSE)
t.test(subset(loan, verification_status == 'Not Verified')$int_rate, subset(loan, verification_status == 'Source Verified')$int_rate, conf.level = 0.95, mu = 0, alternative = "two.sided", paired = FALSE, var.equal = FALSE)
t.test(subset(loan, verification_status == 'Not Verified')$int_rate, subset(loan, verification_status == 'Verified')$int_rate, conf.level = 0.95, mu = 0, alternative = "two.sided", paired = FALSE, var.equal = FALSE)
#pymnt_plan
t.test(subset(loan, pymnt_plan == 'n')$int_rate, subset(loan, pymnt_plan == 'y')$int_rate, conf.level = 0.95, mu = 0, alternative = "two.sided", paired = FALSE, var.equal = FALSE)
#application_type
t.test(subset(loan, application_type == 'INDIVIDUAL')$int_rate, subset(loan, application_type == 'JOINT')$int_rate, conf.level = 0.95, mu = 0, alternative = "two.sided", paired = FALSE, var.equal = FALSE)
#term 
t.test(subset(loan, term == ' 36 months')$int_rate, subset(loan, term == ' 60 months')$int_rate, conf.level = 0.95, mu = 0, alternative = "two.sided", paired = FALSE, var.equal = FALSE)
```
## Statistically significant differences were found among all these four variables: "term", "verification_status", "pymnt_plan", "application_type".


### Step 3 Building Linear Regression Model
## 3.1 Processing selected features/dealing with missing values
```{r}
loan$total_rec_int[which(is.na(loan$total_rec_int))] <- median(loan$total_rec_int, na.rm = T)
loan$revol_util[which(is.na(loan$revol_util))] <- median(loan$revol_util, na.rm = T)
loan$inq_last_6mths[which(is.na(loan$inq_last_6mths))] <- mean(loan$inq_last_6mthsl, na.rm = T)
loan$total_pymnt[which(is.na(loan$total_pymnt))] <- median(loan$total_pymnt, na.rm = T)
loan$total_pymnt_inv[which(is.na(loan$total_pymnt_inv))] <- median(loan$total_pymnt_inv, na.rm = T)
```

## 3.2 Split train and test data set
```{r}
set.seed(1)
train.ind <- sample(1:dim(loan)[1], 0.8 * dim(loan)[1])
train <- loan[train.ind, ]
test <- loan[-train.ind, ]
```

## 3.3 Build model to predict loan$int_rate
```{r}
mod1 <- lm(int_rate ~ total_rec_int + revol_util + inq_last_6mths + total_pymnt + 
             + total_pymnt_inv + term + verification_status + pymnt_plan + application_type, data = train)
summary(mod1)
```
## Given that pymnt_plan is not a significant predictor, it is removed. 
```{r}
mod2 <- lm(int_rate ~ total_rec_int + revol_util + inq_last_6mths + total_pymnt + 
             + total_pymnt_inv + term + verification_status + application_type, data = train)
summary(mod2)
```

## 3.4 Model Performance (accuracy) on test data
```{r}
select.features <- c("int_rate", "total_rec_int", "revol_util", "inq_last_6mths", "total_pymnt","total_pymnt_inv","term","verification_status","application_type")
test.sub <- test[, select.features]
pre <- predict(mod2, newdata = test.sub)
library(hydroGOF)
mse(pre, test.sub$int_rate)
```

```{r}
# plot the model
plot(mod2)
```

## MSE is quite high. This model is not as desired. Let's rebuild the model by adding more features. 

### Step 4 Rebuild the model 
## 4.1 Go back to LC dictionary & more feature engineering
# User Feature Selection
```{r}
#encode home_ownership
loan$home_ownership <- ifelse(loan$home_ownership %in% c('ANY', 'NONE', 'OTHER'), 'OTHER',
                              loan$home_ownership)
#encode state information with the help of int_rate
int_state <- by(loan, loan$addr_state, function(x) {
  return(mean(x$int_rate))
})

loan$state_mean_int <-
  ifelse(loan$addr_state %in% names(int_state)[which(int_state <=
                                                       quantile(int_state, 0.25))], 'low',
         ifelse(loan$addr_state %in% names(int_state)[which(int_state <=
                                                              quantile(int_state, 0.5))],'lowmedium',
                ifelse(loan$addr_state %in% names(int_state)[which(int_state <= quantile(int_state, 0.75))], 
                       'mediumhigh', 'high')))
select.features_1 <- c('home_ownership', 'state_mean_int')

```
# Financial feature selection
```{r}
loan$annual_inc[which(is.na(loan$annual_inc))] <- median(loan$annual_inc, na.rm = T)
select.features_2 <- c('dti', 'annual_inc', 'verification_status')
```
# Credit score feature selection
```{r}
select.features_3 <- c('earliest_cr_line', 'last_credit_pull_d')
```
# Credit line feature selection
```{r}
loan$total_acc[which(is.na(loan$total_acc))] <- mean(loan$total_acc, na.rm = T) 
loan$tot_cur_bal[which(is.na(loan$tot_cur_bal))] <- median(loan$tot_cur_bal, na.rm = T) 
loan$open_acc[which(is.na(loan$open_acc))] <- mean(loan$open_acc, na.rm = T) 
loan$acc_now_delinq[which(is.na(loan$acc_now_delinq))] <- median(loan$acc_now_delinq, na.rm = T)
loan$delinq_2yrs[which(is.na(loan$delinq_2yrs))] <- median(loan$delinq_2yrs, na.rm = T)
loan$mths_since_last_delinq[which(is.na(loan$mths_since_last_delinq))] <- median(loan$mths_since_last_delinq, na.rm = T)
loan$collections_12_mths_ex_med[which(is.na(loan$collections_12_mths_ex_med))] <- median(loan$collections_12_mths_ex_med, na.rm = T)
loan$tot_coll_amt[which(is.na(loan$tot_coll_amt))] <- median(loan$tot_coll_amt, na.rm = T)
loan$pub_rec[which(is.na(loan$pub_rec))] <- median(loan$pub_rec, na.rm = T)
loan$mths_since_last_major_derog[which(is.na(loan$mths_since_last_major_derog))] <- median(loan$mths_since_last_major_derog, na.rm = T)
loan$revol_util[which(is.na(loan$revol_util))] <- mean(loan$revol_util, na.rm = T)
loan$total_rev_hi_lim[which(is.na(loan$total_rev_hi_lim))] <- median(loan$total_rev_hi_lim, na.rm = T)

select.features_4 <- c('total_acc', 'tot_cur_bal', 'open_acc', 'acc_now_delinq', 'delinq_2yrs',
                       'mths_since_last_delinq', 'collections_12_mths_ex_med', 'tot_coll_amt',
                       'pub_rec', 'mths_since_last_major_derog', 'revol_util',
                       'total_rev_hi_lim')
```
# loan feature selection
```{r}
select.features_5 <- c('loan_amnt', 'application_type', 'purpose',
                       'term', 'initial_list_status')
```
# loan payment feature selection
```{r}
select.features_6 <- c('installment', 'funded_amnt', 'funded_amnt_inv', 'pymnt_plan',
                       'recoveries', 'collection_recovery_fee',
                       'out_prncp', 'out_prncp_inv')
```
# Combining all the selected features
```{r}
select.features <- c(select.features_1, select.features_2, select.features_3, select.features_4,
                     select.features_5, select.features_6, 'int_rate')
loan <- loan[select.features]
```
# Engineering numerical & categorical variables
```{r}
# scale numerical features
select.features.num <- names(loan[, sapply(loan[, 1:32], is.numeric)])
loan.scale <- loan
loan.scale[, select.features.num] <- scale(loan.scale[, select.features.num])
# reduce levels of categorical features
select.features.cate <- names(loan.scale[, sapply(loan.scale, is.character)])
n_levels <- sort(sapply(loan.scale[select.features.cate], function(x) {nlevels(as.factor(x))}), decreasing = TRUE)
print(n_levels)
library("zoo")
loan.scale$earliest_cr_line <- format(as.Date(as.yearmon(loan.scale$earliest_cr_line, "%B-%Y")), "%Y")
length(unique(loan.scale$earliest_cr_line))
loan.scale$last_credit_pull_d <- format(as.Date(as.yearmon(loan.scale$last_credit_pull_d, "%B-%Y")), "%Y")
length(unique(loan.scale$last_credit_pull_d))
```

## 4.2 Building new model to predict loan$int_rate
# Split training and test data
```{r}
set.seed(2)
train.ind <- sample(1:dim(loan.scale)[1], 0.8 * dim(loan)[1])
train <- loan.scale[train.ind, ]
test <- loan.scale[-train.ind, ]
```
# Build Linear Regression model
```{r}
mod3 <- lm(int_rate ~ ., data = train)
summary(mod3)
```
## Based on the output, some features are not significant, will then be removed.
```{r}
library(glmnet)
drops <- c("last_credit_pull_d","earliest_cr_line","funded_amnt_inv","pymnt_plan", "int_rate")
ind <- train[, !(names(train) %in% drops)]
ind <- model.matrix( ~., ind)
dep <- train[, 'int_rate']
#Use cross validation to tune parameters
mod4 <- cv.glmnet(ind, dep, family = 'gaussian', alpha = 1.0)
summary(mod4)
plot(mod4)
```
# Optimal parameters for this linear regression model
```{r}
mod4$lambda.1se
coef(mod4, s="lambda.1se")
```

## 4.3 Model Performance on Test Data
```{r}
library(hydroGOF)
ind <- test[, !(names(test) %in% drops)]
ind <- model.matrix( ~., ind)
cv.pred <- predict(mod4, s=mod4$lambda.1se, newx=ind)
mse(cv.pred[,1], test$int_rate)
```

